First Order Alternation

### 2017-05-22 quantifiers/draft.tex
ok

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. Let
$\top$ and $\bot$ be constants of sort $\Bool$.

@@

A \emph{first-order alternating data automaton} (fo-ADA or automaton
in the sequel) is a tuple $\A =
\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
%
\item $G \subseteq \vars$ is a finite set of \emph{global
  variables} of sort $\Data$, 
%
\item $L \subseteq \vars$ is a finite set of \emph{local
  variables} of sort $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset,\emptyset)$ is the \emph{initial
  configuration},
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{G}\cup G\cup\overline{L},L)$
  is a \emph{transition function},
\end{compactitem}
where, for a set $X \subseteq \vars$, we denote $\overline{X} =
\set{\overline{x} \mid x \in X}$. Intuitively, the variables
$\overline{x}$ from a transition formula $\Delta(q,a)$ refer to past
values, whereas the other variables refer to the current
values. Observe that the initial values are unconstrained (since the
initial configuration is a boolean combination of states without
variables) and quantification is allowed only over the current local
variables. In other words, we do not allow quantification over past
values, which are beyond our control. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

@@

Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\phi_0 \Rightarrow \phi_1(Q,\stamp{G}{0} \cup \stamp{G}{1},
\stamp{L}{0} \cup \stamp{L}{1}) \Rightarrow \ldots \Rightarrow
\phi_n(Q,\stamp{G}{0} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{0} \cup
\ldots \cup \stamp{L}{n})\]
where: \begin{compactitem}
\item $\phi_0 \equiv \iota$, 
%
\item $\phi_1 \equiv \exists \stamp{L}{0} ~.~ \stamp{\Delta}{0}(\iota,a_1)$, and
%
\item $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i \in [1,n-1]$.
\end{compactitem}
We say that $\A$ \emph{accepts} $w$ if $\I,\nu \models
\phi_n(Q,\stamp{G}{0} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{0} \cup
\ldots \cup \stamp{L}{n})$, where $\nu$ is a valuation such
that: \begin{compactitem}
\item $\nu(\stamp{g}{i}) = \nu_i(g)$ for each $g \in G$, 
%
\item $\nu(q) = \true$ for each $q \in F$, and
%
\item $\nu(q) = \false$ for each $q \in Q \setminus F$.
\end{compactitem}
The language of $\A$ is the set $L(\A)$ of words from $\Sigma[G]^*$
accepted by $\A$. In the following, we ask two
questions: \begin{compactenum}
\item given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
  with $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 = \emptyset$, do
  there exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such
  that $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[G]^* \setminus L(\A_1)$ ?
%
\item given an automaton $\A$, is $L(\A) = \emptyset$ ? 
\end{compactenum}

>>>>

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
constants of sort $\Bool$ and $\mathbf{0}$ be a constant of sort
$\Data$.

@@

A \emph{first-order alternating data automaton} (fo-ADA or automaton
in the sequel) is a tuple $\A =
\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
%
\item $G \subseteq \vars$ is a finite set of \emph{global
  variables} of sort $\Data$, 
%
\item $L \subseteq \vars$ is a finite set of \emph{local
  variables} of sort $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset,\emptyset)$ is the \emph{initial
  configuration},
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{G}\cup G\cup\overline{L},L)$
  is a \emph{transition function},
\end{compactitem}
where, for a set $X \subseteq \vars$, we denote $\overline{X} =
\set{\overline{x} \mid x \in X}$. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

Intuitively, the variables $\overline{x}$ from a transition formula
$\Delta(q,a)$ refer to past values, whereas the other variables refer
to the current values. Observe that the initial values of the global
variables are unconstrained (since the initial configuration is a
boolean combination of states without variables) and quantification is
allowed only over the current local variables. In other words, we do
not quantify over global variables which are visible throughout the
execution and also, we do not quantify over over past values, which
are beyond our control.

@@

Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\iota \stackrel{(a_1,\nu_1)}{\Longrightarrow}
\phi_1(Q,\stamp{G}{1},
\stamp{L}{1}) \stackrel{(a_2,\nu_2)}{\Longrightarrow} \ldots
\stackrel{(a_n,\nu_n)}{\Longrightarrow} \phi_n(Q,\stamp{G}{1} \cup
\ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup \ldots \cup
\stamp{L}{n})\] where: \begin{compactitem}
\item $\phi_1 \equiv \stamp{\Delta}{0}(\iota,a_1)[\mathbf{0}/\stamp{G}{0},\mathbf{0}/\stamp{L}{0}]$, and
%
\item $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i \in [1,n-1]$.
\end{compactitem}
We say that $\A$ \emph{accepts} $w$ if $\I,\nu \models
\phi_n(Q,\stamp{G}{1} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup
\ldots \cup \stamp{L}{n})$, where $\nu$ is a valuation such
that: \begin{compactitem}
\item $\nu(\stamp{g}{i}) = \nu_i(g)$ for each $g \in G$ and each $i \in [1,n]$, 
%
\item $\nu(q) = \true$ for each $q \in F$, and
%
\item $\nu(q) = \false$ for each $q \in Q \setminus F$.
\end{compactitem}
The language of $\A$ is the set $L(\A)$ of words from $\Sigma[G]^*$
accepted by $\A$. In the following, we ask two
questions: \begin{compactenum}
\item given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, with $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
  \emptyset$, do there exist automata $\A_\cap$, $\A_\cup$ and
  $\overline{\A}_1$ such that $L(\A_\cap) = L(\A_1) \cap L(\A_2)$,
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ and $L(\overline{\A}_1) =
  \Sigma[G]^* \setminus L(\A_1)$ ?
%
\item given an automaton $\A$, is $L(\A) = \emptyset$ ? 
\end{compactenum}
